---
title: "Predicting Train Delays"
author: "Yifei Sun & Kathleen Scopis"
date: "2022-12-16"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: "hide"
    code_download: true
---


# Introduction
### What's all this for?

The New York Metropolitan area has a problem. 

New Jersey Transit (NJT), the third largest public transportation system in the United States, operates a fleet of ~1,200 active trains every day, accounting for early 300 million rides annually across more than 5,000 square miles serviced area.  Featuring rail stations across three states, commuters from across the Northeast rely on the transit network to get to work, school, and other daily activities safely, quickly, and reliably. 

Despite the important role the NJT plays in the lives of its commuters, there is currently no publicly-available system to accurately predict potential train delays in any of the station lines.  This presents a big problem for commuters, as delayed trains drastically affect one's ability to perform daily tasks and arrive to scheduled meetings on time.  Having an accessing which trains are likely to be late puts the power back in the hands of the commuters, who are often left behind in the event of train malfunction. This report aims to suggest one solution to this issue.

This study primarily focuses on the month of January for predicting potential delays across each line serviced by New Jersey Transit.


To view a presentation given regarding this research topic, please visit the following link:
https://www.youtube.com/watch?v=X7f8S3BaEYs


# Data and Methodology 
The following analysis utilizes linear regression, cross validation, and mean absolute error analyses to predict the number of minutes late a train will be, given its station of origin, time of departure, weather conditions, and a temporal lead variable.

Data is pulled from the following link: https://www.kaggle.com/datasets/pranavbadami/nj-transit-amtrak-nec-performance?select=2018_11.csv
The dataset reports on train delay minutes for the entirety of the NJ Transit rail system, as well as selected Amtrak lines, from March 2018 to May 2020. 


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)


library(tidyverse)
library(tigris)
library(sf)
library(chron)
library(timeDate)
library(lubridate)
library(tidycensus)
library(viridis)
library(riem)
library(gridExtra)
library(knitr)
library(kableExtra)
library(RSocrata)
library(ggplot2)
library(spdep)
library(caret)
library(ckanr)
library(grid)
library(ggcorrplot)
library(gganimate)

options(tigris_class = "sf")
source("https://raw.githubusercontent.com/urbanSpatial/Public-Policy-Analytics-Landing/master/functions.r")


plotTheme <- theme(
  plot.title =element_text(size=12),
  plot.subtitle = element_text(size=8),
  plot.caption = element_text(size = 6),
  axis.text.x = element_text(size = 10, angle = 45, hjust = 1),
  axis.text.y = element_text(size = 10),
  axis.title.y = element_text(size = 10),
  # Set the entire chart region to blank
  panel.background=element_blank(),
  plot.background=element_blank(),
  #panel.border=element_rect(colour="#F0F0F0"),
  # Format the grid
  panel.grid.major=element_line(colour="#D0D0D0",size=.2),
  axis.ticks=element_blank())

mapTheme <- theme(plot.title =element_text(size=12),
                  plot.subtitle = element_text(size=8),
                  plot.caption = element_text(size = 6),
                  axis.line=element_blank(),
                  axis.text.x=element_blank(),
                  axis.text.y=element_blank(),
                  axis.ticks=element_blank(),
                  axis.title.x=element_blank(),
                  axis.title.y=element_blank(),
                  panel.background=element_blank(),
                  panel.border=element_blank(),
                  panel.grid.major=element_line(colour = 'transparent'),
                  panel.grid.minor=element_blank(),
                  legend.direction = "vertical", 
                  legend.position = "right",
                  plot.margin = margin(1, 1, 1, 1, 'cm'),
                  legend.key.height = unit(1, "cm"), legend.key.width = unit(0.2, "cm"))

palette5 <- c("#eff3ff","#bdd7e7","#6baed6","#3182bd","#08519c")
palette4 <- c("#D2FBD4","#92BCAB","#527D82","#123F5A")
palette2 <- c("#6baed6","#08519c")


```

# Analysis

### Read Data
First, we read in the data found on Kaggle, as well as the shapefile for the state of New Jersey.  Once imported, we group aspects of the scheduled time of arrival for each train by its expected arrival hour, day, and week.  This will help in later analysis.


```{r read_dat}
dat <- read.csv(file.path("2019_01.csv"))
station <-read.csv(file.path("stops.csv"))
dat<-merge(dat, station, by.x="from", by.y="stop_name")
dat <- dat%>%
  rename(
         from_lat = stop_lat,
         from_lng = stop_lon,
         from_zone_id=zone_id)
dat<-merge(dat, station, by.x="to", by.y="stop_name")
dat <- dat%>%
  rename(
         to_lat = stop_lat,
         to_lng = stop_lon,
         to_zone_id=zone_id)%>%
  dplyr::select(-stop_id.x,-stop_code.x,-stop_id.y,-stop_code.y)
#write.csv(dat, "C:\\Upenn\\d.csv", row.names=FALSE)


#create interval 60 

dat <- dat %>%
  mutate(interval60 = floor_date(ymd_hm(scheduled_time), unit = "hour"),
         week = week(interval60),
         dotw = wday(interval60, label=TRUE))

# load census geography

NJCensus <- 
  get_acs(geography = "tract", 
          variables = c("B01003_001", "B19013_001", 
                        "B02001_002", "B08013_001",
                        "B08012_001", "B08301_001", 
                        "B08301_010", "B01002_001"), 
          year = 2018, 
          state = "NJ", 
          geometry = TRUE, 
          
          output = "wide")

NJTracts <- 
  NJCensus %>%
  as.data.frame() %>%
  distinct(GEOID, .keep_all = TRUE) %>%
  select(GEOID, geometry) %>% 
  st_sf

```

```{r add_census_tracts , message = FALSE, warning = FALSE}
dat_census <- st_join(dat %>% 
          filter(is.na(from_lng) == FALSE &
                   is.na(from_lat) == FALSE &
                   is.na(to_lat) == FALSE &
                   is.na(to_lng) == FALSE) %>%
          st_as_sf(., coords = c("from_lng", "from_lat"), crs = 4326),
        NJTracts %>%
          st_transform(crs=4326),
        join=st_intersects,
              left = TRUE) %>%
  rename(From.Tract = GEOID) %>%
  mutate(from_lng = unlist(map(geometry, 1)),
         from_lat = unlist(map(geometry, 2)))%>%
  as.data.frame() %>%
  select(-geometry)%>%
  st_as_sf(., coords = c("to_lng", "to_lat"), crs = 4326) %>%
  st_join(., NJTracts %>%
            st_transform(crs=4326),
          join=st_intersects,
          left = TRUE) %>%
  rename(To.Tract = GEOID)  %>%
  mutate(to_lng = unlist(map(geometry, 1)),
         to_lat = unlist(map(geometry, 2)))%>%
  as.data.frame() %>%
  select(-geometry)%>%
  na.omit

```


### Evaluating Weather Patterns

To have a better understanding of the potential impact that weather patterns have on train performance, weather data is observed for the NYC weather station from January 2019.  Pulling in information such as temperature, precipitation, wind speed, and overall visibility allows for fluctuations in train driving conditions, as well as may help explain delays caused by inclement weather such as snow and particularly bad storms.

The following graphs highlight an example of the weather conditions later explored in the model: Precipitation, Wind Speed, Temperature, and Visibility.



```{r import_weather, message = FALSE, warning = FALSE}
weather.Data <- 
  riem_measures(station = "WRI", date_start = "2019-01-01", date_end = "2019-01-31")
weather.Panel  <-  
  weather.Data %>%
  dplyr::select(valid, tmpf, p01i, sknt, vsby)%>%
  replace(is.na(.), 0) %>%
    mutate(interval60 = ymd_h(substr(valid,1,13))) %>%
    mutate(week = week(interval60),
           dotw = wday(interval60, label=TRUE)) %>%
    group_by(interval60) %>%
    summarize(Temperature = max(tmpf),
              Precipitation = sum(p01i),
              Wind_Speed = max(sknt),
              Visibility = max(vsby)) %>%
    mutate(Temperature = ifelse(Temperature == 0, 42, Temperature))

#plot
grid.arrange( ncol=2,
  ggplot(weather.Panel, aes(interval60,Precipitation)) + geom_line() + 
    labs(title="Percipitation", x="Hour", y="Perecipitation") + plotTheme,
  ggplot(weather.Panel, aes(interval60,Wind_Speed)) + geom_line() + 
    labs(title="Wind Speed", x="Hour", y="Wind Speed") + plotTheme,
  ggplot(weather.Panel, aes(interval60,Temperature)) + geom_line() + 
    labs(title="Temperature", x="Hour", y="Temperature") + plotTheme,
  ggplot(weather.Panel, aes(interval60,Visibility)) + geom_line() + 
    labs(title="Visibility", x="Hour", y="Visibility") + plotTheme,
  top="Weather Data - NJ Metro Area - 2019")

```


### Further Data Cleaning + Lagged Variables

Further data cleaning is needed in order to render the information useful to building out the model. 

### Commute Patterns by Time

This graph highlights delay patterns across the month of January, as shown in date of scheduled arrival (shown in hour-intervals) and number of delays per hour.  A clear pattern emerges suggesting delays largely happen during peak travel times during the weekday, and are significantly less common during the weekends and off-peak travel hours.


```{r explore data}
ggplot(dat_census %>%
         group_by(interval60) %>%
         tally())+
  geom_line(aes(x = interval60, y=n))+

  labs(title="Delay time per hr. New Jersey, Jan, 2019",
       x="Date", 
       y="Number of minutes")+
  plotTheme


dat_census %>%
        mutate(time_of_day = case_when(hour(interval60) < 7 | hour(interval60) > 18 ~ "Overnight",
                                 hour(interval60) >= 7 & hour(interval60) < 10 ~ "AM Rush",
                                 hour(interval60) >= 10 & hour(interval60) < 15 ~ "Mid-Day",
                                 hour(interval60) >= 15 & hour(interval60) <= 18 ~ "PM Rush"))%>%
         group_by(interval60, from, time_of_day) %>%
         tally()%>%
  group_by(from, time_of_day)%>%
  summarize(mean_delay_minutes = mean(n))%>%
  ggplot()+
  geom_histogram(aes(mean_delay_minutes), binwidth = 1)+
  labs(title="New Jersey, Jan, 2019",
       x="Number of delay minutes", 
       y="Frequency")+
  facet_wrap(~time_of_day)+
  plotTheme
```

The following chart displays the most common delay times seen throughout the NJT system.  Most trains are delayed less than 15 minutes, while a few are delayed more than an hour.  While a 15 minute delay is not necessarily a significant amount of time by itself, it can have severe impacts on a morning or afternoon commute.  Even a few minutes delay can create the need for an unexpected transfer, or a commuter completely missing an important meeting.


```{r}
ggplot(dat_census %>%
         group_by(interval60, line) %>%
         tally())+
  geom_histogram(aes(n), binwidth = 5)+
  labs(title="delay time per hr by train lines. New Jersey, Jan, 2019",
       x="delay time", 
       y="Number of train lines")+
  plotTheme


```


### Commute Patterns by Day of the Week

The following graphs further illustrate average delay patterns seen throughout the week. Delays are significantly more common during the week, particularly during the peak commute hours of 7AM to 9AM, then again from 5PM to 7PM.
This may be due to the increase in rides available during peak hours to accompany increased ride demands, as people crowd onto the platforms to shart their day, then reverse course to return home for the evening. 


```{r}
ggplot(dat_census %>% mutate(hour = hour(actual_time)))+
     geom_freqpoly(aes(hour, color = dotw), binwidth = 1)+
  labs(title="Delay time in New Jersey,  by day of the week, Jan, 2019",
       x="Hour", 
       y="delay time")+
     plotTheme


ggplot(dat_census %>% 
         mutate(hour = hour(actual_time),
                weekend = ifelse(dotw %in% c("Sat", "Sun"), "Weekend", "Weekday")))+
     geom_freqpoly(aes(hour, color = weekend), binwidth = 1)+
  labs(title="Train delay time in New Jersey, - weekend vs weekday, Jan, 2019",
       x="Hour", 
       y="delay time")+
     plotTheme
```

### Spatial Patterns

Analyzing spatial patterns in station delays is a crucial step in understanding underlying causes of delays, as well as potential solutions.  If a station is consistently running under delays, that may signify a deeper problem to the management of the station, overcrowding, or not enough service provided.  

The following maps suggest that stations to the east and center experience the most delays, particularly during the on-peak times.  When combined with the previous graphs, it becomes clear that these stations are simply not recieving enough service during these times, and are likely experiencing a compounding overcrowding problem.

```{r}
ggplot()+
  geom_sf(data = NJTracts %>%
          st_transform(crs=4326))+
  geom_point(data = dat_census %>% 
            mutate(hour = hour(actual_time),
                weekend = ifelse(dotw %in% c("Sat", "Sun"), "Weekend", "Weekday"),
                time_of_day = case_when(hour(interval60) < 7 | hour(interval60) > 18 ~ "Overnight",
                                 hour(interval60) >= 7 & hour(interval60) < 10 ~ "AM Rush",
                                 hour(interval60) >= 10 & hour(interval60) < 15 ~ "Mid-Day",
                                 hour(interval60) >= 15 & hour(interval60) <= 18 ~ "PM Rush"))%>%
              group_by(from, from_lat, from_lng, weekend, time_of_day) %>%
              tally(),
            aes(x=from_lng, y = from_lat, color = n), 
            fill = "transparent", alpha = 1, size = 1.5)+
  scale_colour_viridis(direction = -1,
  discrete = FALSE, option = "D")+
  ylim(min(dat_census$from_lat), max(dat_census$from_lat))+
  xlim(min(dat_census$from_lng), max(dat_census$from_lng))+
  facet_grid(weekend ~ time_of_day)+
  labs(title="Train delay time by station. New Jersey, Jan, 2019")+
  mapTheme
```

### Data Cleaning - Temporal Lead and Holiday Variables

Creating temporally-sensitive variables are essential to creating a robust analysis.  The following code creates variables that observe lead-delay minutes - what was the condition of train delays one minute prior?  What about one day?

Understanding the past conditions is incredibly useful in studying the future, as well as understanding the present. Therefore, the model becomes increasingly more accurate the closer the scheduled time of departure is, as shown in the coefficient table below. 

Moreover, certain days of the year yield much higher ridership, most notably New Year's on January 1st.  The model has been built to address potential holiday demand increase, and therefore a compounding predictor for delay. 

```{r panel_length_check , message = FALSE, warning = FALSE}
length(unique(dat_census$interval60)) * length(unique(dat_census$from_id))
study.panel <- 
  expand.grid(interval60=unique(dat_census$interval60), 
              from_id = unique(dat_census$from_id)) %>%
  left_join(., dat_census %>%
              select(from_id, from, From.Tract, from_lng, from_lat)%>%
              distinct() %>%
              group_by(from_id) %>%
              slice(1))
nrow(study.panel)      

trip.panel <- 
  dat_census %>%
  mutate(Trip_Counter = 1) %>%
  right_join(study.panel) %>% 
  group_by(interval60, from_id, from, From.Tract, from_lng, from_lat) %>%
  summarize(delay_time = sum(Trip_Counter, na.rm=T)) %>%
  left_join(weather.Panel) %>%
  ungroup() %>%
  filter(is.na(from_id) == FALSE) %>%
  mutate(week = week(interval60),
         dotw = wday(interval60, label = TRUE)) %>%
  filter(is.na(From.Tract) == FALSE)


trip_panel<-merge(dat_census, weather.Panel)%>%
  rename(GEOID=From.Tract)%>%
left_join(NJTracts, trip_panel1, by='GEOID')

trip_panel$format_date<-format(trip_panel$interval60, "%Y-%m-%d")

#holidays

trip_panel <- trip_panel %>%
  mutate(holiday = case_when(
    scheduled_time == 1   ~ 1,    # New Year's Day
    scheduled_time == 32  ~ 1,   # Martin Luther King Jr. Day
    scheduled_time == 121 ~ 1,  # Memorial Day
    scheduled_time == 153 ~ 1,  # Independence Day
    scheduled_time == 217 ~ 1,  # Labor Day
    scheduled_time == 286 ~ 1,  # Thanksgiving Day
    scheduled_time == 359 ~ 1,  # Christmas Day
    TRUE                  ~ 0              # Other days
  ))

 
# time lead variables
trip_panel <- 
  trip_panel %>% 
  arrange( interval60) %>% 
  mutate(lagHour = dplyr::lag(delay_minutes,1),
         lag2Hours = dplyr::lag(delay_minutes,2),
         lag3Hours = dplyr::lag(delay_minutes,3),
         lag4Hours = dplyr::lag(delay_minutes,4),
         lag12Hours = dplyr::lag(delay_minutes,12),
         lag1day = dplyr::lead(delay_minutes,24),
         leadHour = dplyr::lead(delay_minutes,1),
         lead2Hours = dplyr::lead(delay_minutes,2),
         lead3Hours = dplyr::lead(delay_minutes,3),
         lead4Hours = dplyr::lead(delay_minutes,4),
         lead12Hours = dplyr::lead(delay_minutes,12),
         lead1day = dplyr::lead(delay_minutes,24)
        )%>%
  mutate(holidayLag = case_when(dplyr::lag(holiday, 1) == 1 ~ "PlusOneDay",
                                 dplyr::lag(holiday, 2) == 1 ~ "PlustTwoDays",
                                 dplyr::lag(holiday, 3) == 1 ~ "PlustThreeDays",
                                 dplyr::lead(holiday, 1) == 1 ~ "MinusOneDay",
                                 dplyr::lead(holiday, 2) == 1 ~ "MinusTwoDays",
                                 dplyr::lead(holiday, 3) == 1 ~ "MinusThreeDays"))%>%
    
   mutate(holidayLag = ifelse(is.na(holidayLag),0,holidayLag))
        
    
  

# correlation


as.data.frame(trip_panel) %>%
    group_by(interval60) %>% 
    summarise_at(vars(starts_with("lead"), "delay_minutes"), mean, na.rm = TRUE) %>%
    gather(Variable, Value, -interval60, -delay_minutes) %>%
    mutate(Variable = factor(Variable, levels=c("leadHour","lead2Hours","lead3Hours","lead1day")))%>%
    group_by(Variable) %>%  
    summarize(correlation = round(cor(Value, delay_minutes),2))
```




### Data Cleaning - Test vs. Training

The following step is crucial to predicting delays.  First, the dataset is split into a training set, in which the algorithm becomes familiar with the trends in the numbers according to the variables it is being fed, as well as a test set, which will lay out the actual predictions.

Three linear regression models are built and then compared against each other. 
Reg1 only includes the basic weather variables as detailed above, while the second 

As we only have data of only one month, we use data of week1 to week3 to be the training set, and week4 & week5 to be the test set, due to the data is very large, the R-square is very small. However, the MAE is also small, the MAE of the best regression results is 2.93, which means the average of prediction and real value is 2.93 minutes. And that shows our model has a good accuracy

```{r}
trip.Train <- filter(trip_panel, week < 4)
trip.Test <- filter(trip_panel, week >= 4)


mondays <- 
  mutate(trip_panel,
         monday = ifelse(dotw == "周一" & hour(interval60) == 1,
                         interval60, 0)) %>%
  filter(monday != 0) 

tg   <- as.POSIXct("2018-11-22 01:00:00 UTC")
xmas <- as.POSIXct("2018-12-24 01:00:00 UTC")

st_drop_geometry(rbind(
  mutate(trip.Train, Legend = "Training"), 
  mutate(trip.Test, Legend = "Testing"))) %>%
    group_by(Legend, interval60) %>% 
      summarize(delay_minutes = sum(delay_minutes)) %>%
      ungroup() %>% 
      ggplot(aes(interval60, delay_minutes, colour = Legend)) + geom_line() +
        scale_colour_manual(values = palette2) +
        geom_vline(data = mondays, aes(xintercept = monday)) +
        labs(title="delay time by week: 12.31-1.28",
              
             x="Day", y="Trip Count") +
         theme(panel.grid.major = element_blank())   




# Build regression models

reg1 <- lm(delay_minutes~Temperature+Precipitation+Wind_Speed+Visibility, data=trip.Train )
reg2 <- 
  lm(delay_minutes ~  line+holiday +  hour(interval60) + Temperature + Precipitation +Wind_Speed+Visibility+from, 
     data=trip.Train)
reg3 <- 
  lm(delay_minutes ~  line+holiday +  hour(interval60) + Temperature + Precipitation +Wind_Speed+Visibility+leadHour+lead2Hours+lead3Hours+lead4Hours+lead12Hours+lead1day+from, 
     data=trip.Train)

```
        

```{r}
trip.Test.weekNest <- 
  as.data.frame(trip.Test) %>%
  nest(-week) 


model_pred <- function(dat, fit){
   pred <- predict(fit, newdata = dat)}
```


### Mean Absolute Errors


Mean Absolute Errors (MAE) are calculated by finding the average distance between predicted and observed values.  A smaller MAE represents a better fit model and can help in determining which variables are more important.
Overall, the areas to the center and east have higher MAEs.  This may be explained by the fact that these stations also have higher delay minutes, and are subjected to potentially larger errors.

```{r do_predicitons }
week_predictions <- 
  trip.Test.weekNest %>% 
    mutate(reg1= map(.x = data, fit = reg1, .f = model_pred),
           reg2= map(.x = data, fit = reg2, .f = model_pred),
           reg3= map(.x = data, fit = reg3, .f = model_pred),) %>% 
    gather(Regression, Prediction, -data, -week) %>%
    mutate(Observed = map(data, pull, delay_minutes),
           Absolute_Error = map2(Observed, Prediction,  ~ abs(.x - .y)),
           MAE = map_dbl(Absolute_Error, mean, na.rm = TRUE),
           sd_AE = map_dbl(Absolute_Error, sd, na.rm = TRUE))
week_predictions


plot1<-week_predictions %>%
  dplyr::select(week, Regression, MAE) %>%
  gather(Variable, MAE, -Regression, -week) %>%
  ggplot(aes(week, MAE)) + 
    geom_bar(aes(fill = Regression), position = "dodge", stat="identity") +
    scale_fill_manual(values = palette5) +
    labs(title = "Mean Absolute Errors by model specification and week") 
 plot(plot1)


week_predictions %>% 
    mutate(interval60 = map(data, pull, interval60),
           from_id = map(data, pull, from_id), 
           from_lat = map(data, pull, from_lat), 
           from_lng = map(data, pull, from_lng)) %>%
    select(interval60, from_id, from_lng, from_lat, Observed, Prediction, Regression) %>%
    unnest() %>%
  filter(Regression == "reg3") %>%
  group_by(from_id, from_lng, from_lat) %>%
  summarize(MAE = mean(abs(Observed-Prediction), na.rm = TRUE))%>%
ggplot(.)+
  geom_sf(data = NJCensus, color = "black", fill = "transparent")+
  geom_point(aes(x = from_lng, y = from_lat, color = MAE), 
             fill = "transparent", alpha = 0.7,size=3)+
  scale_colour_viridis(direction = -1,
  discrete = FALSE, option = "D")+
  ylim(min(dat_census$from_lat), max(dat_census$from_lat))+
  xlim(min(dat_census$from_lng), max(dat_census$from_lng))+
  labs(title="Mean Abs Error, Test Set, Model 3")+
  mapTheme
```
### Linear Regression

The following line graph shows the results of the three linear regression models.  While all three perform relatively well, the third, which includes all predictor variables, is a fantastic match and is well fit to the dataset. 


```{r error_vs_actual_timeseries , warning = FALSE, message = FALSE}
week_predictions %>% 
    mutate(interval60 = map(data, pull, interval60),
           from_id = map(data, pull, from_id)) %>%
    dplyr::select(interval60, from_id, Observed, Prediction, Regression) %>%
    unnest() %>%
    gather(Variable, Value, -Regression, -interval60, -from_id) %>%
    group_by(Regression, Variable, interval60) %>%
    summarize(Value = sum(Value)) %>%
    ggplot(aes(interval60, Value, colour=Variable)) + 
      geom_line(size = 1.1) + 
      facet_wrap(~Regression, ncol=1) +
      labs(title = "Predicted/Observed delay time in time series", subtitle = "NJ; A test set of 2 weeks",  x = "Hour", y= "delay minutes") +
      plotTheme
```


We also checked the error of prediction result in time and space 
The first plot shows we underpredict the delay time, especially during the weekends and rush hour.
The second plot shows that MAE’s distribution in space has a certain pattern, but not very clear. During the weekdays, the prediction is not very precise in the AM rush and PM rush, but perform well in Mid-day and night. During the weekend, the prediction not perform well in mid-day.
In conclusion, the goodness of our prediction shows some pattern in time and space, which is the limitation of our model, we should use more suitable feature to fit the model, or use GWR model to solve the spatial deviation.

```{R}
week_predictions %>% 
    mutate(interval60 = map(data, pull, interval60),
           from_id = map(data, pull, from_id), 
           from_lat = map(data, pull, from_lat), 
           from_lng = map(data, pull, from_lng),
           dotw = map(data, pull, dotw)) %>%
    select(interval60, from_id, from_lng, 
           from_lat, Observed, Prediction, Regression,
           dotw) %>%
    unnest() %>%
  filter(Regression == "reg3")%>%
  mutate(weekend = ifelse(dotw %in% c("Sat", "Sun"), "Weekend", "Weekday"),
         time_of_day = case_when(hour(interval60) < 7 | hour(interval60) > 18 ~ "Overnight",
                                 hour(interval60) >= 7 & hour(interval60) < 10 ~ "AM Rush",
                                 hour(interval60) >= 10 & hour(interval60) < 15 ~ "Mid-Day",
                                 hour(interval60) >= 15 & hour(interval60) <= 18 ~ "PM Rush"))%>%
  ggplot()+
  geom_point(aes(x= Observed, y = Prediction))+
    geom_smooth(aes(x= Observed, y= Prediction), method = "lm", se = FALSE, color = "red")+
    geom_abline(slope = 1, intercept = 0)+
  facet_grid(time_of_day~weekend)+
  labs(title="Observed vs Predicted",
       x="Observed trips", 
       y="Predicted trips")+
  plotTheme
```

The following charts observe MAE across temporally-oriented variables. The AM rush has significantly higher MAE, which may be explained by the increase in ridership coming from the suburbs. 



```{r}
week_predictions %>% 
    mutate(interval60 = map(data, pull, interval60),
           from_id = map(data, pull, from_id), 
           from_lat = map(data, pull, from_lat), 
           from_lng = map(data, pull, from_lng),
           dotw = map(data, pull, dotw) ) %>%
    select(interval60, from_id, from_lng, 
           from_lat, Observed, Prediction, Regression,
           dotw) %>%
    unnest() %>%
  filter(Regression == "reg3")%>%
  mutate(weekend = ifelse(dotw %in% c("Sun", "Sat"), "Weekend", "Weekday"),
         time_of_day = case_when(hour(interval60) < 7 | hour(interval60) > 18 ~ "Overnight",
                                 hour(interval60) >= 7 & hour(interval60) < 10 ~ "AM Rush",
                                 hour(interval60) >= 10 & hour(interval60) < 15 ~ "Mid-Day",
                                 hour(interval60) >= 15 & hour(interval60) <= 18 ~ "PM Rush")) %>%
  group_by(from_id, weekend, time_of_day, from_lng, from_lat) %>%
  summarize(MAE = mean(abs(Observed-Prediction), na.rm = TRUE))%>%
  ggplot(.)+
  geom_sf(data = NJCensus, color = "grey", fill = "transparent")+
  geom_point(aes(x = from_lng, y = from_lat, color = MAE), 
             fill = "transparent", size = 1.3, alpha = 0.8)+
  scale_colour_viridis(direction = -1,
  discrete = FALSE, option = "D")+
  ylim(min(dat_census$from_lat), max(dat_census$from_lat))+
  xlim(min(dat_census$from_lng), max(dat_census$from_lng))+
  facet_grid(weekend~time_of_day)+
  labs(title="Mean Absolute Errors, Test Set")+
  mapTheme
```

### Cross validation
To test the model’s generalization, we use K-folds to do the cross validation, and set k=30, the results shows that the MAE is 3.13 and the RMSE is 5.3, which shows a good generalization of our model.


```{r cross validation}
fitControl <- trainControl(method = "cv", number = 30)
set.seed(825)

reg.cv <- 
  train(delay_minutes ~ ., data = trip_panel %>% 
                                dplyr::select(delay_minutes,line,holiday ,Temperature, Precipitation,Wind_Speed,Visibility,lagHour ,lag2Hours ,lag3Hours , lag12Hours , lag1day , holidayLag,leadHour,lead2Hours,lead3Hours,lead4Hours,lead12Hours,lead1day,from), 
     method = "lm", trControl = fitControl, na.action = na.pass)

reg.cv


reg.cv$resample[1:5,]
```



# Conclusion + Suggestions

After analyzing the results of our analysis, it is clear that this model may be recommended for use by daily commuters on all lines of the New Jersey Transit system. 



# Final Thoughts + App Wireframe

Due to the proven accuracy of the model, we recommend New Jersey Transit adopt the model for use in predicting delays in the Northeast Corridor.  

[https://www.figma.com/proto/63yCCvuvYQjDE8CXtZTpHd/Untitled?node-id=69%3A488&scaling=scale-down&page-id=0%3A1&starting-point-node-id=13%3A121]




